Time for a rethink: time sub-sampling methods in disparity-through-time analyses

### 2018-01-30 manuscript/response_to_reviewers.tex

\section{Reviewer 1 (Steve Brusatte)}

\begin{enumerate}

\item{\textcolor{blue}{In the section on adding in reconstructed ancestors to the dataset before ordination, I suppose you should mention that this is similar to the approach that we outlined in our Paleobiology paper (Brusatte et al. 2011) and mention any key differences.}}

We now explicitly refer to Brusatte et al. 2011 and mention the key difference:

\textit{This approach is similar to \cite{brusatte2011phylogenetic} but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).} lines @@@

\item{\textcolor{blue}{I’m not sure about the use of the term ‘cladisto-spaces’. To me this implies that the space itself is ordinated using phylogenetic information (e.g., branch lengths). Now, you are indeed including some phylogenetic information in your morphospaces, in terms of reconstructed ancestors and inferred states for branches. So sure, I think what you have could be called a ‘cladisto-space’. Same with most of the morphospace options produced by Claddis. But the ones that most of us have done in the past, simply using discrete characters to ordinate a morphospace, haven’t generally incorporated phylogeny. I’m talking about the references you cite in lines 141-143 here. These morphospaces been ordinated using discrete characters of the type that are used in phylogenetic analysis. But the characters are not phylogenetically transformed in any way. As far as the ordination analysis knows, they are simply discrete character statements. So I wonder if the term ‘cladisto-space’ is confusing, and is just yet another new term that may muddy the waters rather than clarify. If you want to keep using this term, that’s fine, but a sentence or two of explanation (about how previous authors have used discrete characters in a non-phylogenetic way) would be warranted.}}

%TG: let's just remove cladisto-space: I think (1) it's a nice word and (2) it's actually a good decoy to focus reviewers attention and can be changed easily. + Steve makes a point with adding to many jargon.
%TG: actually I had a go at trying to change it and this might require your writing skills to make it flow (+ I'm hangover). I've tried a paragraph (commented out) in the relevant section.

@@

%TG: good idea, we (I) can make a figure designating the three different strategies.

\item{\textcolor{blue}{Line 158. Gower’s distance is not the Euclidean distance (= the square root of the squared distances) over the number of shared characters. For a set of unordered characters, it is the number of mismatched character states over the number of comparable/shared characters. For a set of ordered characters, it is sum of the differences in the values of the characters states for each pairwise comparison, over the number of comparable characters. This is an edit that needs to be made only to the text; because the authors used functions from the Claddis package (see github documentation), they would have calculated Gower’s distance correctly.}}

\item{\textcolor{blue}{Consider using ``stage'' instead of ``age'' when referring to the finer resolution time scale.}}

We've changed the ``stratigraphic age'' to ``stratigraphic stage'' throughout the manuscript.

### 2018-01-30 manuscript/time-slice.tex
Updated response to review

\paragraph{Estimating ancestral character states.}
\label{ace}
For each dataset we estimated the ancestral character states at each node using the \texttt{AncStatesEstMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis,R}.
This function uses the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny \citep[based on the \texttt{rerootingMethod} function in \texttt{phytools};][]{phytools}.
Inapplicable and missing characters for any taxon were treated as ambiguous characters (i.e. any possible observed state for the character).
% NC: I've left this one bit to make it clear but removed the explanation (this was all just copied from your thesis)
To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').
This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state reconstruction methods, especially in parts of the datasets where data were sparse.

@@

\paragraph{Constructing distance matrices.}
To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For cladistic matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see Supporting Information Appendix S1 for details of which species).

@@

\begin{enumerate}
  \item Stratigraphic time bins. This is the traditional method, where all the taxa within each stratigraphic period are included in the disparity calculation. This often leads to bins of unequal duration. Here we use stratigraphic ages and epochs.
  \item Equally sized time bins. This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation. 
  \item Time-slicing. We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation. 
\end{enumerate}

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
    \caption[Relative disparity-through-time.]
    {Relative disparity-through-time. Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, blue points are time-slices with a punctuated model of evolution (`proximity' method), and green points are time-slices with a gradual model of evolution (`gradual splits' method).
    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods. 
    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
    The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A1-A2.}
    \label{figure:dtt1}
\end{figure}

@@

% Plots showing the differences
\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Timing of peak disparity.
    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, green points are time-slices with a punctuated model of evolution (`proximity' method), and blue points are time-slices with a gradual model of evolution (`gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table \ref{table:datasets} for details).
    Results for stratigraphic ages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A3-A4.}
    \label{figure:peak1}
\end{figure}

@@

% Big heat map table thing
\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-extinction-effect.pdf}
    \caption[Extinction example datasets.]
    {Effects of mass extinction events on disparity.
    Pink cells and blue cells indicate respectively a significant or non significant change in disparity before and after the mass extinction event (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455-430 MYA; \citealt{wright2017bayesian}).
    \textit{e:1}, \textit{e:2}, and \textit{e:3} denote whether the comparison was between the time bin or time-slice immediately after the mass extinction (\textit{e:1}), or the second (\textit{e:2}) or third (\textit{e:3}) bin/slice after the mass extinction to account for any lag effect.
    The top seven rows use stratigraphic ages and the bottom seven rows use stratigraphic epochs.
    Labels on the left hand side indicate whether time bins (``bins'') were used, or which of the six time-slicing methods was used.}
    \label{figure:extinction1}
\end{figure}

@@

Time-slicing has several advantages over time binning (using either equally or unequally sized bins) approaches.
First, it allows us to use as much of the information available to us, in the form of phylogenetic relationships and ancestral taxa, as possible. 
This increases our ability to investigate key biological questions, such as how do various drivers influence morphological diversity through time, and how do mass extinctions influence disparity \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}, both by increasing the statistical power of analyses and through the availability of data at key time points in the history of our groups.
Second, we are able to be more explicit about the mode of evolution in our clades; in time-slicing we can apply punctuated or gradual models of trait change rather than making an assumption of punctuated evolution.
This may be important, as gradual change is a common pattern of trait evolution in the fossil record \citep{Hunt20112007}.
Of course the method also has limitations.
The main one of these is a practical one; it requires a time-calibrated phylogeny.
Not all palaeontological datasets have phylogenies yet.
Another limitation is that the method, like most phylogeny based methods, depends on ancestral state reconstructions.
Care should be taken in interpreting these, as they are highly dependent on the data and models used for the reconstructions \citep{Slateretal2012,doi:10.1080/10635150801910451}.

>>>>

\paragraph{Estimating ancestral character states.}
\label{ace}
For each dataset we estimated the ancestral character states at each node using the \texttt{AncStatesEstMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis,R}.
This function uses the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny \citep[based on the \texttt{rerootingMethod} function in \texttt{phytools};][]{phytools}.
Inapplicable and missing characters for any taxon were treated as ambiguous characters (i.e. any possible observed state for the character).
% NC: I've left this one bit to make it clear but removed the explanation (this was all just copied from your thesis)
To prevent poor ancestral state estimations from biasing our results, especially when a lot of error is associated with the estimations, we only included ancestral state estimations with a scaled Likelihood $\geq$ $0.95$.
Ancestral state estimations with scaled Likelihoods below this threshold were recoded as missing data (``?'').
This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state estimations methods, especially in parts of the datasets where data were sparse. 
This approach is similar to \cite{brusatte2011phylogenetic} but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).

@@

% To explore disparity-through-time in our datasets we used a discrete morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
% This approach is a approximation of a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010} and is based on discrete morphological data (i.e. the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
% Although empirically inter-taxon distances are the same in a morphospace based on discrete or continuous data \citep{foth2012different,hetherington2015cladistic}, we like to highlight that this space is based on discrete morphological data and phylogeny (it could be referred to as a ``cladisto-space'').
% Mathematically, a cladisto-space is an $n$ dimensional object that summarises the distances between the taxa present in a discrete morphological matrix.

\paragraph{Constructing distance matrices.}
To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For discrete morphological matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see Supporting Information Appendix S1 for details of which species).

@@

\begin{enumerate}
  \item Stratigraphic time bins. This is the traditional method, where all the taxa within each stratigraphic period are included in the disparity calculation. This often leads to bins of unequal duration. Here we use stratigraphic stages and epochs.
  \item Equally sized time bins. This is another commonly used method, where the time frame of interest is split into equally sized time bins, then all the taxa within each time bin are included in the disparity calculation. 
  \item Time-slicing. We describe this in more detail below, but in brief, time-slicing uses a phylogeny, and rather than binning the data, it takes slices through a phylogeny and includes all the taxa and nodes in that slice within the disparity calculation. 
\end{enumerate}

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
    \caption[Relative disparity-through-time.]
    {Relative disparity-through-time. Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, blue points are time-slices with a punctuated model of evolution (`proximity' method), and green points are time-slices with a gradual model of evolution (`gradual splits' method).
    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods. 
    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
    The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
    Results for stratigraphic stages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A1-A2.}
    \label{figure:dtt1}
\end{figure}

@@

% Plots showing the differences
\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Timing of peak disparity.
    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, green points are time-slices with a punctuated model of evolution (`proximity' method), and blue points are time-slices with a gradual model of evolution (`gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (see Table \ref{table:datasets} for details).
    Results for stratigraphic stages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A3-A4.}
    \label{figure:peak1}
\end{figure}

@@

% Big heat map table thing
\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-extinction-effect.pdf}
    \caption[Extinction example datasets.]
    {Effects of mass extinction events on disparity.
    Pink cells and blue cells indicate respectively a significant or non significant change in disparity before and after the mass extinction event (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455-430 MYA; \citealt{wright2017bayesian}).
    \textit{e:1}, \textit{e:2}, and \textit{e:3} denote whether the comparison was between the time bin or time-slice immediately after the mass extinction (\textit{e:1}), or the second (\textit{e:2}) or third (\textit{e:3}) bin/slice after the mass extinction to account for any lag effect.
    The top seven rows use stratigraphic stages and the bottom seven rows use stratigraphic epochs.
    Labels on the left hand side indicate whether time bins (``bins'') were used, or which of the six time-slicing methods was used.}
    \label{figure:extinction1}
\end{figure}

@@

Time-slicing has several advantages over time binning (using either equally or unequally sized bins) approaches.
First, it allows us to use as much of the information available to us, in the form of phylogenetic relationships and ancestral taxa, as possible. 
This increases our ability to investigate key biological questions, such as how do various drivers influence morphological diversity through time, and how do mass extinctions influence disparity \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}, both by increasing the statistical power of analyses and through the availability of data at key time points in the history of our groups.
Second, we are able to be more explicit about the mode of evolution in our clades; in time-slicing we can apply punctuated or gradual models of trait change rather than making an assumption of punctuated evolution.
This may be important, as gradual change is a common pattern of trait evolution in the fossil record \citep{Hunt20112007}.
Of course the method also has limitations.
The main one of these is a practical one; it requires a time-calibrated phylogeny.
Not all palaeontological datasets have phylogenies yet.
Another limitation is that the method, like most phylogeny based methods, depends on ancestral state estimations.
Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
